---
title: "r_exploration_notes"
author: "Judith Bourque"
date: "`r Sys.Date()`"
output: md_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Description

A collection of notes and resources for personal use.

# R

## R file types

- R script `.R` - analysis mode and want a report as a side effect
- R Markdown `.Rmd` - writing a report with a lot of R code in it
- R Notebook - R Markdown document with chunks that can be executed independently and interactively, with output visible immediately beneauth the input.

# R Markdown

Resources

- [R Markdown: The Definitive Guide](https://bookdown.org/yihui/rmarkdown/)
- [R Markdown from RStudio](https://rmarkdown.rstudio.com/lesson-1.html)
- [Pandoc User's Guide](https://pandoc.org/MANUAL.html)
- [R Markdown Reference guide](https://www.rstudio.com/wp-content/uploads/2015/03/rmarkdown-reference.pdf?_ga=2.164587583.1800627063.1648159135-864680125.1644240072)
- [R Markdown Formats](https://rmarkdown.rstudio.com/formats.html)
- [R Markdown Cookbook](https://bookdown.org/yihui/rmarkdown-cookbook/)
- [bookdown: Authoring Books and Technical Documents with R Markdown](https://bookdown.org/yihui/bookdown/)
- [Bookdown](https://github.com/rstudio/bookdown)
- [Citation Management and Writing Tools: LaTeX and BibTex](https://libguides.mit.edu/cite-write/bibtex)
- [RMarkdown how to](http://www.ece.ualberta.ca/~terheide/rmarkdown-how-to/markdown.html#more_information)

## YAML

Optional header surrounded by `---`

```{r eval = FALSE}
---
title: "Title of my Document"
subtitle: "Subtitle of my Document"
author: "Judith Bourque"
# Date updates automatically
date: '`r Sys.Date()`'
output: 
  github_document:
  # Table of contents
    toc: true
    number_sections: true
# Parameters set values automatically
params:
  data: "data"
---
```

When you put output as github_document, you will knit a .md. Useful as it is a GitHub friendly markdown document and quite pleasant to look at.

Parameters are values you can set when you render the report. They help to re-render the same report with distinct values for various key inputs.

To access a parameter in code, call `params$<parameter name>`.

## Output formats

Documents

- `github_document` - GitHub Flavored Markdown document
- `md_document`
- `pdf_document` - PDF document (via LaTeX template)
- `word_document` Mirosoft Word document (docx)
- `context_document`
- `html_document`
- `latex_document`
- `odt_document` - OpenDocument Text document
- `rtf_document`

Interactive documents

- `htmlwidgets` package
- `shiny` package
- `html_notebook` - Interactive R Notebooks

Presentations

- `powerpoint_presentation`
- `beamer_presentation` - PDF presentation with LaTeX Beamer
- `slidy_presentation`

Other

- `flexdashboard::flex_dashboard` - interactive dashboards


Resource

- [Output formats](https://rmarkdown.rstudio.com/lesson-9.html)

## Render output

If you don't want to knit, run the `render` command

```{r eval = FALSE}
library(rmarkdown)
render("example.Rmd")
```

## Headers

```{md}
# Numbered header
# Unnumbered header {-}
# Also unnumbered header {.unnumbered}
```

## Blockquotes

Code

```{md}
> "This is a quote that can be long and go until another line."
>
> --- Author
```

Output

> "This is a quote that can be long and go until another line."
>
> --- Author
## Markdown Syntax Basics

Pandoc's Markdown

- Images - `![alt text or image title](path/to/image)`
- Footnotes - `^[This is a footnote.]`
- Horizontal rules - three or more of *, -, _

Resource

- [Pandoc User's Guide](https://pandoc.org/MANUAL.html)

## Citations

BibTeX databases - plain-texte file with filename extension `.bib`

```{r eval = FALSE}
@Manual{R-base,
  title = {R: A Language and Environment for Statistical
    Computing},
  author = {{R Core Team}},
  organization = {R Foundation for Statistical Computing},
  address = {Vienna, Austria},
  year = {2017},
  url = {https://www.R-project.org/},
}
```

## Tables

Display options for data frames and matrixes

- Default - like in the R terminal
- `knitr::kable` - pretty

```{r eval = FALSE}
library(knitr)
kable(mtcars[1:5, ], caption = "A knitr kable.")
```

## Code chunks

Code chunks can be run individually in RStudio by clicking the green arrow pointing right. 

Code chunks start with ```` ```{}```` and ends with ```` ``` ````.

```{r eval = FALSE}
# Code displayed without running
# How to put a code chunk in a code chunk?
``` 

Code chunk options

- `include = FALSE` - prevents code and results from appearing in the output file. R Markdown still runs the code in the chunk, and the results can be used by other chunks.
- `echo = FALSE` -  prevents code, but not the results from appearing in the finished file. Good to embed figures.
- `message = FALSE` - prevents messages that are generated by code from appearing in the finished file.
- `warning = FALSE` - prevents warnings that are generated by code from appearing in the finished.
- `fig.cap = "..."` - adds a caption to graphical results.
- `eval = FALSE`

Global options

To set global options that apply to every chunk in your file, call `knitr::opts_chunk$set` in a code chunk. Knitr will treat each option that you pass to `knitr::opts_chunk$set` as a global default that can be overwritten in individual chunk headers.

## Inline code

Insert code results directly by enclosing the code with `` `r ` ``

Example

`There are 10 individuals.` becomes ``There are `nrow(my_data)` individuals.``

To include *n* literal backticks, use at least *n + 1* backticks outside, e.g., you can use four backticks to preserve three backtick inside ````` ```` ```code``` ```` `````, which is rendered as ```` ```code``` ````.

# R for data science

Resources

- [R for Data Science](https://r4ds.had.co.nz/index.html)
- [R Programming for Data Science](https://bookdown.org/rdpeng/rprogdatascience/)
- [R for Data Science: Exercice Solutions](https://jrnold.github.io/r4ds-exercise-solutions/)
- [R for Graduate Students](https://bookdown.org/yih_huynh/Guide-to-R-Book/)
- [thesisdown](https://github.com/ismayc/thesisdown)

## Tidyverse package

`library(tidyverse)`

`tidyverse` is a collection of R packages designed for data science.

Core packages loaded automatically with `library(tidyverse)`

- `ggplot2` - create graphics
- `dplyr` - manipulate data
- `tidyr` - tidy data
- `readr` - read rectangular data (csv, tsv)
- `purrr` - loops
- `tibble` - data frame 2.0, they do less and complain more forcing you to adress problems earlier
- `stringr` - functions to work with strings
- `forcats` - tools to work with factors

Resources

- [Tidyverse](https://www.tidyverse.org/)
- [The Tidyverse Cookbook](https://rstudio-education.github.io/tidyverse-cookbook/index.html)

# Resources

- [Introduction to R](http://r.sund.ku.dk/index.html)
- [RStudio Cheatsheets](https://www.rstudio.com/resources/cheatsheets/)
- [Using R and RStudio for Data Management, Statistical Analysis, and Graphics](https://englianhu.files.wordpress.com/2016/01/using-r-and-rstudio-for-data-management-statistical-analysis-and-graphics-2nd-edit.pdf)
- [ProgrammingR](https://www.programmingr.com/)
- [Advanced R](https://adv-r.hadley.nz/index.html)

Course documents

- [STAT 545 Data wrangling, exploration, and analysis with R](https://stat545.com/index.html)
- [Happy Git and GitHub for the useR: Installation](https://happygitwithr.com/install-intro.html)

Data analysis

- [Exploratory Data Analysis](https://mgimond.github.io/ES218/Week01.html)
- [Geocomputation with R](https://geocompr.robinlovelace.net/) - geographic data analysis

Data Visualization

- [Data Visualization with R](https://rkabacoff.github.io/datavis/)
- [Data Visualization](https://socviz.co/index.html#preface)
- [Fundamentals of Data Visualization](https://clauswilke.com/dataviz/)

Text mining

- [Welcome to Text Mining with R](https://www.tidytextmining.com/index.html)
- [Notes for “Text Mining with R: A Tidy Approach”](https://bookdown.org/Maxine/tidy-text-mining/)
- [Work with strings with stringr cheat sheet](https://evoldyn.gitlab.io/evomics-2018/ref-sheets/R_strings.pdf)
- [Stringr](https://stringr.tidyverse.org/) (see cheat sheet for regex)

Databases

- [Databases using R](https://db.rstudio.com/getting-started/)
- [10 Easy Steps to a Complete Understanding of SQL](https://blog.jooq.org/10-easy-steps-to-a-complete-understanding-of-sql/)
- [SQL Query Planning](https://www.sqlite.org/queryplanner.html)

Art

- [Creating Generative Art with R](https://cpastem.caltech.edu/documents/18024/ARt.nb.html)
- [Tiny Art with R (R studio)](https://alinastepanova.medium.com/tiny-art-with-r-d6d93d110619)
- [Tweetable Mathematical Art With R](https://fronkonstin.com/2018/09/06/tweetable-mathematical-art-with-r/)

Miscellaneous

- [What Every Programmer Absolutely, Positively Needs To Know About Encodings And Character Sets To Work With Text](https://kunststube.net/encoding/)
- [making memes in R](http://jenrichmond.rbind.io/post/making-memes-in-r/)
- [WikipediR](https://github.com/Ironholds/WikipediR)
- [Learn to purrr](https://www.rebeccabarter.com/blog/2019-08-19_purrr/#:~:text=Purrr%20is%20one%20of%20those,purrr%20is%20all%20about%20iteration.)
- [Statistical Inference via Data Science](https://moderndive.com/)
- [Introduction to Data Science](https://rafalab.github.io/dsbook/)
- [Efficient R programming](https://csgillespie.github.io/efficientR/)
- [Engineering Production-Grade Shiny Apps](https://engineering-shiny.org/)
- [R Packages](https://r-pkgs.org/index.html)
- [Reader expectations approach](https://georgegopen.com/litigation-articles)
