---
title: "r-exploration notes"
author: "Judith Bourque"
date: "`r Sys.Date()`"
output: md_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Description

A collection of quick notes and resources for personal use.

# R basics

```{r echo = FALSE}
df <- data.frame(
  name = c(
    "R script",
    "R Markdown",
    "R Notebook"),
  purpose = c(
    "analysis mode and want a report as a side effect",
    "writing a report with a lot of R code in it",
    "R Markdown document with chunks that can be executed independently and interactively, with output visible immediately beneauth the input"),
  extension = c(
    ".R",
    ".Rmd",
    ""))
knitr::kable(df, caption = "R file types")
```

Resources

-   [Introduction to R](http://r.sund.ku.dk/index.html)
-   [RStudio Cheatsheets](https://www.rstudio.com/resources/cheatsheets/)
-   [Using R and RStudio for Data Management, Statistical Analysis, and Graphics](https://englianhu.files.wordpress.com/2016/01/using-r-and-rstudio-for-data-management-statistical-analysis-and-graphics-2nd-edit.pdf)
-   [ProgrammingR](https://www.programmingr.com/)
-   [Advanced R](https://adv-r.hadley.nz/index.html)
-   [Efficient R programming](https://csgillespie.github.io/efficientR/)

## Workflow

- GitHub
- Renv

Resources

-   [Happy Git and GitHub for the useR: Installation](https://happygitwithr.com/install-intro.html) (course)

# R Markdown

Resources

-   [R Markdown: The Definitive Guide](https://bookdown.org/yihui/rmarkdown/)
-   [R Markdown from RStudio](https://rmarkdown.rstudio.com/lesson-1.html)
-   [R Markdown Reference guide](https://www.rstudio.com/wp-content/uploads/2015/03/rmarkdown-reference.pdf?_ga=2.164587583.1800627063.1648159135-864680125.1644240072)
-   [R Markdown Formats](https://rmarkdown.rstudio.com/formats.html)
-   [R Markdown Cookbook](https://bookdown.org/yihui/rmarkdown-cookbook/)
-   [RMarkdown how to](http://www.ece.ualberta.ca/~terheide/rmarkdown-how-to/markdown.html#more_information)
-   [bookdown: Authoring Books and Technical Documents with R Markdown](https://bookdown.org/yihui/bookdown/)
-   [Bookdown](https://github.com/rstudio/bookdown)

## YAML

Optional header surrounded by `---`

```{r eval = FALSE}
---
title: "Title of my Document"
subtitle: "Subtitle of my Document"
author: "Judith Bourque"
# Date updates automatically
date: '`r Sys.Date()`'
output: 
  github_document:
  # Table of contents
    toc: true
    number_sections: true
# Parameters set values automatically
params:
  data: "data"
---
```

When you put output as github_document, you will knit a .md. Useful as it is a GitHub friendly markdown document and quite pleasant to look at.

Parameters are values you can set when you render the report. They help to re-render the same report with distinct values for various key inputs.

To access a parameter in code, call `params$<parameter name>`.

### Output formats

Documents

-   `github_document`: GitHub Flavored Markdown document
-   `md_document`
-   `pdf_document`: PDF document (via LaTeX template)
-   `word_document` Mirosoft Word document (docx)
-   `context_document`
-   `html_document`
-   `latex_document`
-   `odt_document`: OpenDocument Text document
-   `rtf_document`

Interactive documents

-   `htmlwidgets`: package
-   `shiny`: package
-   `html_notebook`: Interactive R Notebooks

Presentations

-   `powerpoint_presentation`
-   `beamer_presentation`: PDF presentation with LaTeX Beamer
-   `slidy_presentation`

Other

-   `flexdashboard::flex_dashboard`: interactive dashboards

Resource

-   [Output formats](https://rmarkdown.rstudio.com/lesson-9.html)

### Render output

If you don't want to knit, run the `render` command

```{r eval = FALSE}
library(rmarkdown)
render("example.Rmd")
```

## Headers

```{md}
# Numbered header
# Unnumbered header {-}
# Also unnumbered header {.unnumbered}
```

## Markdown Syntax

```{r echo = FALSE}
df <- data.frame(
  syntax = c(
    "Images",
    "Footnotes",
    "Horizontal rules"),
  code = c(
    "`![alt text or image title](path/to/image)`",
    "`^[This is a footnote.]`",
    "three or more of *, -, _"))
knitr::kable(df, caption = "Pandoc's Markdown")
```

Resource

-   [Pandoc User's Guide](https://pandoc.org/MANUAL.html)

## Tables

Display options for data frames and matrixes

-   Default: like in the R terminal
-   `knitr::kable`: pretty

Code for creating a table using a new data frame

```{r eval = FALSE}
# ```{r echo = FALSE}
df <- data.frame(
  column1 = c(
    "1",
    "2",
    "3",
    "4",
    "5"),
  column2 = c(
    "1",
    "2",
    "3",
    "4",
    "5"))
knitr::kable(df, caption = "caption")
# ```
```

## Code chunks

Code chunks can be run individually in RStudio by clicking the green arrow pointing right.

Code chunks start with ```` ```{} ```` and ends with ```` ``` ````.

```{r eval = FALSE}
# ```{r}
# ```
```

```{r echo = FALSE}
df <- data.frame(
  option = c(
    "`include = FALSE`",
    "`echo = FALSE`",
    "`message = FALSE`",
    "`warning = FALSE`",
    "`fig.cap = \"...\"`",
    "`eval = FALSE`"),
  purpose = c(
    "prevents code and results from appearing in the output file. R Markdown still runs the code in the chunk, and the results can be used by other chunks",
    "prevents code, but not the results from appearing in the finished file. Good to embed figures",
    "prevents messages that are generated by code from appearing in the finished file",
    "prevents warnings that are generated by code from appearing in the finished",
    "adds a caption to graphical results",
    ""))
knitr::kable(df, caption = "Code chunk options")
```

Global options

To set global options that apply to every chunk in your file, call `knitr::opts_chunk$set` in a code chunk. Knitr will treat each option that you pass to `knitr::opts_chunk$set` as a global default that can be overwritten in individual chunk headers.

## Inline code

Insert code results directly by enclosing the code with `` `r ` ``

Example: `There are 10 individuals.` becomes `` There are `nrow(my_data)` individuals. ``

To include *n* literal backticks, use at least *n + 1* backticks outside, e.g., you can use four backticks to preserve three backtick inside ````` ```` ```code``` ```` `````, which is rendered as ```` ```code``` ````.

## Citations

BibTeX databases: plain-texte file with filename extension `.bib`. Tool and a file format which are used to describe and process lists of references, mostly in conjunction with LaTeX documents.

```{r eval = FALSE}
@Manual{R-base,
  title = {R: A Language and Environment for Statistical
    Computing},
  author = {{R Core Team}},
  organization = {R Foundation for Statistical Computing},
  address = {Vienna, Austria},
  year = {2017},
  url = {https://www.R-project.org/},
}
```

Resources

-   [Citation Management and Writing Tools: LaTeX and BibTex](https://libguides.mit.edu/cite-write/bibtex)
-   [BibTex](http://www.bibtex.org/)
-   [JabRef](https://www.jabref.org/#): open-source reference manager compatible with BibTex. Works on Windows.

## Blockquotes

Code

```{md}
> "This is a quote that can be long and go until another line."
>
> --- Author
```

Output

> "This is a quote that can be long and go until another line."
>
> --- Author

# R for data science

Resources

-   [R for Data Science](https://r4ds.had.co.nz/index.html)
-   [R Programming for Data Science](https://bookdown.org/rdpeng/rprogdatascience/)
-   [R for Data Science: Exercice Solutions](https://jrnold.github.io/r4ds-exercise-solutions/)
-   [R for Graduate Students](https://bookdown.org/yih_huynh/Guide-to-R-Book/)
-   [thesisdown](https://github.com/ismayc/thesisdown)
-   [Statistical Inference via Data Science](https://moderndive.com/)
-   [Introduction to Data Science](https://rafalab.github.io/dsbook/)
-   [STAT 545 Data wrangling, exploration, and analysis with R](https://stat545.com/index.html) (course)

## Tidyverse package

`tidyverse` is a collection of R packages designed for data science.

Core packages loaded automatically with `library(tidyverse)`

```{r echo = FALSE}
df <- data.frame(
  package = c("ggplot2", "dplyr", "tidyr", "readr", "purrr","tibble", "stringr", "forcats"),
  purpose = c("create graphics", "manipulate data", "tidy data", "read rectangular data (csv, tsv)", "loops", "data frame 2.0, they do less and complain more forcing you to adress problems earlier", "functions to work with strings", "tools to work with factors"))
knitr::kable(df, caption = "tidyverse core packages")
```

Resources

-   [Tidyverse](https://www.tidyverse.org/)
-   [The Tidyverse Cookbook](https://rstudio-education.github.io/tidyverse-cookbook/index.html)
-   [Learn to purrr](https://www.rebeccabarter.com/blog/2019-08-19_purrr/#:~:text=Purrr%20is%20one%20of%20those,purrr%20is%20all%20about%20iteration.)

## Data import

Resources

-   [Databases using R](https://db.rstudio.com/getting-started/)
-   [10 Easy Steps to a Complete Understanding of SQL](https://blog.jooq.org/10-easy-steps-to-a-complete-understanding-of-sql/)
-   [SQL Query Planning](https://www.sqlite.org/queryplanner.html)

## Data analysis

Resources

-   [Exploratory Data Analysis](https://mgimond.github.io/ES218/Week01.html)
-   [Geocomputation with R](https://geocompr.robinlovelace.net/): geographic data analysis

## Text mining

-   [Welcome to Text Mining with R](https://www.tidytextmining.com/index.html)
-   [Notes for "Text Mining with R: A Tidy Approach"](https://bookdown.org/Maxine/tidy-text-mining/)
-   [Work with strings with stringr cheat sheet](https://evoldyn.gitlab.io/evomics-2018/ref-sheets/R_strings.pdf)
-   [Stringr](https://stringr.tidyverse.org/) (see cheat sheet for regex)
-   [What Every Programmer Absolutely, Positively Needs To Know About Encodings And Character Sets To Work With Text](https://kunststube.net/encoding/)

## Data visualization

- ggplot2: plot
- gganimate: animate plot
- ggrepel: repel labels from overlap
- ggridges: create ridges

Resources

-   [Data Visualization with R](https://rkabacoff.github.io/datavis/)
-   [Data Visualization](https://socviz.co/index.html#preface)
-   [Fundamentals of Data Visualization](https://clauswilke.com/dataviz/)
- [ggridges](https://wilkelab.org/ggridges/index.html)

### Interactive graphs

- leaflet: interactive maps
- plotly: make any ggplot2 interactive
- rCharts: switch graph format with one click (ex: grouped vs. stacked barchart)
- highcharter: show and hide graph elements (and option to download graph)

# Other resources

Art

-   [Creating Generative Art with R](https://cpastem.caltech.edu/documents/18024/ARt.nb.html)
-   [Tiny Art with R (R studio)](https://alinastepanova.medium.com/tiny-art-with-r-d6d93d110619)
-   [Tweetable Mathematical Art With R](https://fronkonstin.com/2018/09/06/tweetable-mathematical-art-with-r/)
-   [making memes in R](http://jenrichmond.rbind.io/post/making-memes-in-r/)

Other

-   [Engineering Production-Grade Shiny Apps](https://engineering-shiny.org/)
-   [R Packages](https://r-pkgs.org/index.html)
-   [WikipediR](https://github.com/Ironholds/WikipediR)
-   [Reader expectations approach](https://georgegopen.com/litigation-articles)
