---
title: "r-exploration notes"
author: "Judith Bourque"
date: "`r Sys.Date()`"
output: md_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Description

A collection of quick notes and References for personal use.

# R basics

```{r echo = FALSE}
df <- data.frame(
  name = c(
    "R script",
    "R Markdown",
    "R Notebook"),
  purpose = c(
    "analysis mode and want a report as a side effect",
    "writing a report with a lot of R code in it",
    "R Markdown document with chunks that can be executed independently and interactively, with output visible immediately beneauth the input"),
  extension = c(
    ".R",
    ".Rmd",
    ""))
knitr::kable(df, caption = "R file types")
```

References

-   [Introduction to R](http://r.sund.ku.dk/index.html)
-   [RStudio Cheatsheets](https://www.rstudio.com/resources/cheatsheets/)
-   [Using R and RStudio for Data Management, Statistical Analysis, and Graphics](https://englianhu.files.wordpress.com/2016/01/using-r-and-rstudio-for-data-management-statistical-analysis-and-graphics-2nd-edit.pdf)
-   [ProgrammingR](https://www.programmingr.com/)
-   [Advanced R](https://adv-r.hadley.nz/index.html)
-   [Efficient R programming](https://csgillespie.github.io/efficientR/)

# Workflow

Programs

-   R
-   RStudio
-   GitHub: git repository
-   GitBash: command line

New project setup

1.  Create and clone repository from GitHub
2.  Create .Rproject
3.  Use `renv::init()` to initialize renv with a new or existing project. Enter autorisations needed.
4.  Set up .gitignore

File and folder structure

-   /code: contains .R files
-   /graphs: contains graphs
-   README.md: describe the project

# Renv

Your project may make use of packages which are available from remote sources requiring some form of authentication to access. Usually, either a personal access token (PAT) or username + password combination is required for authentication. renv is able to authenticate when downloading from such sources.

Credentials can be stored in .Renviron, or can be set in your R session through other means as appropriate.

If you require custom authentication for different packages (for example, your project makes use of packages available on different GitHub enterprise servers), you can use the renv.auth R option to provide package-specific authentication settings. renv.auth can either be a a named list associating package names with environment variables, or a function accepting a package name + record, and returning a list of environment variables. For example:

```{}
# define a function providing authentication
options(renv.auth = function(package, record) {
      if (package == "MyPackage")
        return(list(GITHUB_PAT = "<pat>"))
    })

# use a named list directly
    options(renv.auth = list(
      MyPackage = list(GITHUB_PAT = "<pat>")
    ))

    # alternatively, set package-specific option
    options(renv.auth.MyPackage = list(GITHUB_PAT = "<pat>"))
```

## .Rprofile

.Rprofile files are user-controllable files to set options and environment variables. `.Rprofile` files can be either at the user or project level. User-level .Rprofile files live in the base of the user's home directory, and project-level .Rprofile files live in the base of the project directory. 

R will source only one .Rprofile file. So if you have both a project-specific .Rprofile file and a user .Rprofile file that you want to use, you explicitly source the user-level .Rprofile at the top of your project-level .Rprofile with source("~/.Rprofile").

.Rprofile files are sourced as regular R code, so setting environment variables must be done inside a Sys.setenv(key = "value") call. 

One easy way to edit your .Rprofile file is to use the usethis::edit_r_profile() function from within an R session. You can specify whether you want to edit the user or project level .Rprofile.

## .Renviron

.Renviron is a user-controllable file that can be used to create environment variables. This is especially useful to avoid including credentials like API keys inside R scripts. This file is written in a key-value format, so environment variables are created in the format:

```{}
Key1=value1
Key2=value2
```

And then Sys.getenv("Key1") will return "value1" in an R session.

Like with the .Rprofile file, .Renviron files can be at either the user or project level. If there is a project-level .Renviron, the user-level file will not be sourced. The usethis package includes a helper function for editing .Renviron files from an R session with `usethis::edit_r_environ()`.

References

-   [renv](https://rstudio.github.io/renv/articles/renv.html#authentication): documentation
- [Managing R with .Rprofile, .Renviron, Rprofile.site, Renviron.site, rsession.conf, and repos.conf](https://support.rstudio.com/hc/en-us/articles/360047157094-Managing-R-with-Rprofile-Renviron-Rprofile-site-Renviron-site-rsession-conf-and-repos-conf#:~:text=Rprofile%20files%20are%20user%2Dcontrollable,base%20of%20the%20project%20directory.)

# Git and GitHub

```{r echo = FALSE}
df <- data.frame(
  command = c(
    "cd",
    "cd ..",
    "git status",
    "git diff",
    "git add",
    "git add -A",
    "git commit -m \"\"",
    "git push",
    "git pull",
    "git rm --cached"),
  purpose = c(
    "",
    "",
    "View project changes",
    "View difference in file",
    "",
    "Add all files",
    "",
    "",
    "",
    "stop tracking a file that is currently tracked"))
knitr::kable(df, caption = "git commands")
```

References

-   [Happy Git and GitHub for the useR: Installation](https://happygitwithr.com/install-intro.html): course

## gitignore

A `.gitignore` file is a plain text file where each line contains a pattern for files/directories to ignore. Generally, this is placed in the root folder of the repository, and that's what I recommend. However, you can put it in any folder in the repository and you can also have multiple `.gitignore` files. The patterns in the files are relative to the location of that `.gitignore` file.

```{r echo = FALSE}
df <- data.frame(
  code = c(
    "*",
    "!",
    "#"),
  purpose = c(
    "Matches 0 or more characters, except /",
    "Not ignore a file that would be ignored",
    "Comments"))
knitr::kable(df, caption = ".gitignore code")
```

`.gitignore` options

    # Private files
    _*

To ignore some files in all repositories on your computer, put them in a global `.gitignore` file. First, you have to add a setting to Git with this command:

    git config --global core.excludesFile ~/.gitignore

To ignore a currently tracked file, use this command:

    git rm --cached FILENAME

References

-   [gitignore](https://git-scm.com/docs/gitignore): official documentation

## GitHub Flavored Markdown

References

- [Writing on GitHub](https://docs.github.com/en/get-started/writing-on-github): tips for GitHub flavored markdown and other writing formats
- [Github Flavored Markdown cheatsheet](https://gist.github.com/stevenyap/7038119)

# R Markdown

References

-   [R Markdown: The Definitive Guide](https://bookdown.org/yihui/rmarkdown/)
-   [R Markdown from RStudio](https://rmarkdown.rstudio.com/lesson-1.html)
-   [R Markdown Reference guide](https://www.rstudio.com/wp-content/uploads/2015/03/rmarkdown-reference.pdf?_ga=2.164587583.1800627063.1648159135-864680125.1644240072)
-   [R Markdown Formats](https://rmarkdown.rstudio.com/formats.html)
-   [R Markdown Cookbook](https://bookdown.org/yihui/rmarkdown-cookbook/)
-   [RMarkdown how to](http://www.ece.ualberta.ca/~terheide/rmarkdown-how-to/markdown.html#more_information)
-   [bookdown: Authoring Books and Technical Documents with R Markdown](https://bookdown.org/yihui/bookdown/)
-   [Bookdown](https://github.com/rstudio/bookdown)

## YAML

Optional header surrounded by `---`

```{r eval = FALSE}
---
title: "Title of my Document"
subtitle: "Subtitle of my Document"
author: "Judith Bourque"
# Date updates automatically
date: '`r Sys.Date()`'
output: 
  github_document:
  # Table of contents
    toc: true
    number_sections: true
# Parameters set values automatically
params:
  data: "data"
---
```

When you put output as github_document, you will knit a .md. Useful as it is a GitHub friendly markdown document and quite pleasant to look at.

Parameters are values you can set when you render the report. They help to re-render the same report with distinct values for various key inputs.

To access a parameter in code, call `params$<parameter name>`.

### Output formats

Documents

-   `github_document`: GitHub Flavored Markdown document
-   `md_document`
-   `pdf_document`: PDF document (via LaTeX template)
-   `word_document` Mirosoft Word document (docx)
-   `context_document`
-   `html_document`
-   `latex_document`
-   `odt_document`: OpenDocument Text document
-   `rtf_document`

Interactive documents

-   `htmlwidgets`: package
-   `shiny`: package
-   `html_notebook`: Interactive R Notebooks

Presentations

-   `powerpoint_presentation`
-   `beamer_presentation`: PDF presentation with LaTeX Beamer
-   `slidy_presentation`

Other

-   `flexdashboard::flex_dashboard`: interactive dashboards

Resource

-   [Output formats](https://rmarkdown.rstudio.com/lesson-9.html)

### Render output

If you don't want to knit, run the `render` command

```{r eval = FALSE}
library(rmarkdown)
render("example.Rmd")
```

## Headers

```{md}
# Numbered header
# Unnumbered header {-}
# Also unnumbered header {.unnumbered}
```

## Markdown Syntax

```{r echo = FALSE}
df <- data.frame(
  syntax = c(
    "Images",
    "Footnotes",
    "Horizontal rules"),
  code = c(
    "`![alt text or image title](path/to/image)`",
    "`^[This is a footnote.]`",
    "three or more of *, -, _"))
knitr::kable(df, caption = "Pandoc's Markdown")
```

Resource

-   [Pandoc User's Guide](https://pandoc.org/MANUAL.html)

## Tables

Display options for data frames and matrixes

-   Default: like in the R terminal
-   `knitr::kable`: pretty

Code for creating a table using a new data frame

```{r eval = FALSE}
# ```{r echo = FALSE}
df <- data.frame(
  column1 = c(
    "1",
    "2",
    "3",
    "4",
    "5"),
  column2 = c(
    "1",
    "2",
    "3",
    "4",
    "5"))
knitr::kable(df, caption = "caption")
# ```
```

## Code chunks

Code chunks can be run individually in RStudio by clicking the green arrow pointing right.

Code chunks start with ```` ```{} ```` and ends with ```` ``` ````.

```{r eval = FALSE}
# ```{r}
# ```
```

```{r echo = FALSE}
df <- data.frame(
  option = c(
    "`include = FALSE`",
    "`echo = FALSE`",
    "`message = FALSE`",
    "`warning = FALSE`",
    "`fig.cap = \"...\"`",
    "`eval = FALSE`"),
  purpose = c(
    "prevents code and results from appearing in the output file. R Markdown still runs the code in the chunk, and the results can be used by other chunks",
    "prevents code, but not the results from appearing in the finished file. Good to embed figures",
    "prevents messages that are generated by code from appearing in the finished file",
    "prevents warnings that are generated by code from appearing in the finished",
    "adds a caption to graphical results",
    ""))
knitr::kable(df, caption = "Code chunk options")
```

Global options

To set global options that apply to every chunk in your file, call `knitr::opts_chunk$set` in a code chunk. Knitr will treat each option that you pass to `knitr::opts_chunk$set` as a global default that can be overwritten in individual chunk headers.

## Inline code

Insert code results directly by enclosing the code with `` `r ` ``

Example: `There are 10 individuals.` becomes `` There are `nrow(my_data)` individuals. ``

To include *n* literal backticks, use at least *n + 1* backticks outside, e.g., you can use four backticks to preserve three backtick inside ````` ```` ```code``` ```` `````, which is rendered as ```` ```code``` ````.

## Citations

BibTeX databases: plain-texte file with filename extension `.bib`. Tool and a file format which are used to describe and process lists of references, mostly in conjunction with LaTeX documents.

```{r eval = FALSE}
@Manual{R-base,
  title = {R: A Language and Environment for Statistical
    Computing},
  author = {{R Core Team}},
  organization = {R Foundation for Statistical Computing},
  address = {Vienna, Austria},
  year = {2017},
  url = {https://www.R-project.org/},
}
```

References

-   [Citation Management and Writing Tools: LaTeX and BibTex](https://libguides.mit.edu/cite-write/bibtex)
-   [BibTex](http://www.bibtex.org/)
-   [JabRef](https://www.jabref.org/#): open-source reference manager compatible with BibTex. Works on Windows.

## Blockquotes

Code

```{md}
> "This is a quote that can be long and go until another line."
>
> --- Author
```

Output

> "This is a quote that can be long and go until another line."
>
> --- Author

# R for data science

References

-   [R for Data Science](https://r4ds.had.co.nz/index.html)
-   [R Programming for Data Science](https://bookdown.org/rdpeng/rprogdatascience/)
-   [R for Data Science: Exercice Solutions](https://jrnold.github.io/r4ds-exercise-solutions/)
-   [R for Graduate Students](https://bookdown.org/yih_huynh/Guide-to-R-Book/)
-   [thesisdown](https://github.com/ismayc/thesisdown)
-   [Statistical Inference via Data Science](https://moderndive.com/)
-   [Introduction to Data Science](https://rafalab.github.io/dsbook/)
-   [STAT 545 Data wrangling, exploration, and analysis with R](https://stat545.com/index.html) (course)

## Tidyverse package

`tidyverse` is a collection of R packages designed for data science.

Core packages loaded automatically with `library(tidyverse)`

```{r echo = FALSE}
df <- data.frame(
  package = c("ggplot2", "dplyr", "tidyr", "readr", "purrr","tibble", "stringr", "forcats"),
  purpose = c("create graphics", "manipulate data", "tidy data", "read rectangular data (csv, tsv)", "loops", "data frame 2.0, they do less and complain more forcing you to adress problems earlier", "functions to work with strings", "tools to work with factors"))
knitr::kable(df, caption = "tidyverse core packages")
```

References

-   [Tidyverse](https://www.tidyverse.org/)
-   [The Tidyverse Cookbook](https://rstudio-education.github.io/tidyverse-cookbook/index.html)
-   [Learn to purrr](https://www.rebeccabarter.com/blog/2019-08-19_purrr/#:~:text=Purrr%20is%20one%20of%20those,purrr%20is%20all%20about%20iteration.)
-   [Forcats](https://forcats.tidyverse.org/reference/fct_reorder.html)

# Data import

Workflow

-   If authentification is required to access the data, update `.Renviron` file using `usethis::edit_r_environ()`

References

-   [Databases using R](https://db.rstudio.com/getting-started/)
-   [10 Easy Steps to a Complete Understanding of SQL](https://blog.jooq.org/10-easy-steps-to-a-complete-understanding-of-sql/)
-   [SQL Query Planning](https://www.sqlite.org/queryplanner.html)

## Wikimedia packages

- [WikipediR](https://cran.r-project.org/web/packages/WikipediR/WikipediR.pdf): Wrapper for MediaWiki API
- [WikidataR](https://cran.r-project.org/web/packages/WikidataR/index.html): read and query wikidata. Uses Wikidata and Quickstatements APIs.
- [pageviews](https://cran.r-project.org/web/packages/pageviews/pageviews.pdf): retrieve wikimedia pageviews
- [wmfdata](https://github.com/wikimedia/wmfdata-r): for working with Wikimedia data in R
- [waxer](https://github.com/wikimedia/waxer): querying Wikimedia Analytics Query Service in R.
- [wmfastr](https://github.com/wikimedia/wmfastr): Speedily computes various dwelltime and preference-related metrics
  in the context of search sessions.

# Data analysis

References

-   [Exploratory Data Analysis](https://mgimond.github.io/ES218/Week01.html)
-   [Geocomputation with R](https://geocompr.robinlovelace.net/): geographic data analysis

# Text mining

-   [Welcome to Text Mining with R](https://www.tidytextmining.com/index.html)
-   [Notes for "Text Mining with R: A Tidy Approach"](https://bookdown.org/Maxine/tidy-text-mining/)
-   [Work with strings with stringr cheat sheet](https://evoldyn.gitlab.io/evomics-2018/ref-sheets/R_strings.pdf)
-   [Stringr](https://stringr.tidyverse.org/) (see cheat sheet for regex)
-   [What Every Programmer Absolutely, Positively Needs To Know About Encodings And Character Sets To Work With Text](https://kunststube.net/encoding/)

# Data visualization

```{r echo = FALSE}
df <- data.frame(
  package = c(
    "ggplot2",
    "gganimate",
    "ggrepel",
    "ggridges",
    "viridis"),
  purpose = c(
    "plot",
    "animate plot",
    "repel labels from overlap",
    "create ridges",
    "colour-blind friendly color maps for R"))
knitr::kable(df, caption = "data visualization packages")
```

References

- [ggplot2: elegant graphics for data analysis](https://ggplot2-book.org/index.html)
- [R graphics cookbook](https://r-graphics.org/)
-   [Data Visualization with R](https://rkabacoff.github.io/datavis/)
-   [Data Visualization](https://socviz.co/index.html#preface)
-   [Fundamentals of Data Visualization](https://clauswilke.com/dataviz/)
-   [R graph gallery](https://r-graph-gallery.com/index.html): overview of R graphs and code
-   [ggridges](https://wilkelab.org/ggridges/index.html)
-   [from Data to Vis](https://www.data-to-viz.com/): choose data visualization format based on data type
- [strategies for avoiding the spaghetti graph](https://www.storytellingwithdata.com/blog/2013/03/avoiding-spaghetti-graph): what not to do with line graphs
- [GGplot cheatsheet](https://github.com/rstudio/cheatsheets/blob/main/data-visualization-2.1.pdf)
- [ggrepel](https://ggrepel.slowkow.com/articles/examples.html)
- [patchwork](https://patchwork.data-imaginist.com/): combine ggplots

## Interactive graphs

-   leaflet: interactive maps
-   plotly: make any ggplot2 interactive
-   rCharts: switch graph format with one click (ex: grouped vs. stacked barchart)
-   highcharter: show and hide graph elements (and option to download graph)

## Data art

-   [Creating Generative Art with R](https://cpastem.caltech.edu/documents/18024/ARt.nb.html)
-   [Tiny Art with R (R studio)](https://alinastepanova.medium.com/tiny-art-with-r-d6d93d110619)
-   [Tweetable Mathematical Art With R](https://fronkonstin.com/2018/09/06/tweetable-mathematical-art-with-r/)
-   [making memes in R](http://jenrichmond.rbind.io/post/making-memes-in-r/)

# Other References

Other

-   [Engineering Production-Grade Shiny Apps](https://engineering-shiny.org/)
-   [R Packages](https://r-pkgs.org/index.html)
-   [WikipediR](https://github.com/Ironholds/WikipediR)
-   [Reader expectations approach](https://georgegopen.com/litigation-articles)
